Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-23 MS_thesis.tex
Minor changes to acknowledgements

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis. Historically, much of this foundation has been built using models of infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in a world where population sizes  routinely fluctuate. In this thesis, I analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how trait frequencies change over time. These SDEs reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and may reverse the direction of evolution predicted by infinite-population frameworks. My SDEs also recover well-known results such as the replicator-mutator equation and the Price equation in the infinite population limit, and help explain and organize several previous studies under a single theoretical framework. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations. I also discuss the implications for designing evolutionary agent-based models and empirical studies.
}

@@

I am also grateful to various non-human companions, canine, feline, and more exotic. The dogs and cats of ISI are more helpful to my mental health than I could possibly articulate. Even though my thesis work may seem dreadfully abstract, running after pretty animals in the Western Ghats and other wild places of India is how I even got interested in science in the first place, and I certainly wouldn't be here today if the animals weren't so fascinating :)

>>>>

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
\textcolor{blue}{Several recent theoretical studies have shown that noise can have strong impacts on evolutionary dynamics in the limit of small population sizes. In this thesis, I analytically describe the evolutionary dynamics of finite fluctuating populations from first principles to capture the fundamental phenomena underlying such noise-induced effects. Starting from a density-dependent `birth-death process’ describing a population of individuals with discrete traits, I derive stochastic differential equations (SDEs) for how the relative population sizes and trait frequencies change over time. These SDEs generically reveal a directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and is present even when all types have the same fitness.  The strength of noise-induced selection depends directly on the difference in turnover rates between types and inversely on the total population size. Noise-induced selection can reverse the direction of evolution predicted by infinite-population frameworks. This general derivation of evolutionary dynamics helps unify and organize several previous studies -- typically performed for specific evolutionary and ecological contexts -- under a single set of equations. My SDEs also recover well-known results such as the replicator-mutator equation, the Price equation, and Fisher’s fundamental theorem in the infinite population limit, illustrating consistency with known formal descriptions of evolution. Finally, I extend the birth-death formalism to one-dimensional quantitative traits through a `stochastic field theory’ that yields equations such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit and provides an alternative approach to modelling the evolution of quantitative traits that is more accessible than current measure-theoretic approaches.}
}

@@

I am also grateful to various non-human companions, canine, feline, and more exotic. The dogs and cats around my home are more helpful to my mental health than I could possibly articulate. Even though my thesis work may seem dreadfully abstract, running after animals in the Western Ghats and other wild places of India is how I got interested in science in the first place, and I certainly wouldn't be here today if the animals weren't so fascinating :)
